Open archaeology, open source?

### 2023-10-12 analysis/_05-collaboration.qmd
Reorder section 5

Users can also engage with repositories and other users with social media-like features such as *starring* a repository, adding it to a public or private list, or *following* a user.
These actions populate a timeline of through which users can see recent activity and discover new projects related to those they have interacted with in the past.^[This feature of GitHub's timeline was one of the primary ways we compiled open-archaeo.]
While not constituting a direct form of collaboration, these features can facilitate the formation and maintenance of collaborative networks, in the same way that other social media platforms serve other professional networks.

```{r fig-lifespan}
#| fig-cap: Lifespan of repositories based on the date of first and last commit. Points indicate intermediate commits. Only repositories with more than one commit are included.
#TODO: This looks cool, but is it actually more useful than a histogram or two?
oarch %>%
  unnest(gh_commits) %>%
  drop_na(datetime) %>%
  group_by(item_name) %>%
  add_tally() %>%
  filter(n > 1) %>%
  mutate(
    first_commit = min(datetime, na.rm = TRUE),
    mid_commit = median(datetime, na.rm = TRUE),
    last_commit = max(datetime, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  # geom_linerange(size = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(limits = c(ymd("2005-01-01", tz = "UTC"),
                              ymd("2021-05-31", tz = "UTC")),
                   date_breaks = "2 years", date_labels = "%Y") +
  # coord_flip() +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

@@

The lifespan and activity (commit rate) of repositories in open-archaeo vary greatly (@fig-lifespan).
Here we refer to the *development* lifespan of a project—measured here as the time between a repository's first and last commit—which is not necessary linked to its use-life.
The average project lasts <!-- TODO --> days with <!-- TODO --> commits per day.
<!-- TODO: turn all these percentages into code -->
Many projects are active for only a short period of time: about 15% less than a month, 25% less than three months, and 40% less than a year.
The longest are <!-- TODO -->.
Commit rates vary between <!-- TODO -->, with peaks at <!-- TODO -->.

@@

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\]. A "contribution", in this case, is a commit, issue, or comment applied by a user to a repository.

**NOTE:** Re-calculate these numbers to ensure they are up to date. Translate the first point to be about how many repos have a quarter of contributions by a single person, so that it corresponds with the vertical lines on the graphic, which are now divided into quarters rather than thirds.

```{r plot-contrib-dist}
oarch %>%
  unnest(gh_contribs) %>%
  drop_na(contributor) %>%
  group_by(item_name) %>%
  mutate(n_contributors = n()) %>%
  filter(n_contributors > 1) %>%
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) %>% 
  group_by(item_name, big_boy) %>%
  summarise(contributions = sum(contributions)) %>%
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) %>%
  ungroup() %>%
  mutate(item_name = fct_reorder(item_name, p_big_boy)) %>% 
  ggplot(aes(x = item_name, y = contributions,
             fill = fct_relevel(big_boy, c("Other contributors",
                                           "Leading contributor")))) +
  geom_col(position = "fill", colour = "white", size = 0.1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast(guide = guide_legend(reverse = TRUE)) +
  labs(title = "Repository contributors by proportion of contributions",
       x = NULL, y = "Contributions",
       fill = NULL,
       caption = "Data: open-archaeo.info and GitHub API") +
  coord_flip() +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

@@

>>>>

```{r fig-lifespan}
#| fig-cap: Lifespan of repositories based on the date of first and last commit. Points indicate intermediate commits. Only repositories with more than one commit are included.
#TODO: This looks cool, but is it actually more useful than a histogram or two?
#TODO: Remove that one commit in 2001
oarch %>%
  unnest(gh_commits) %>%
  drop_na(datetime) %>%
  group_by(item_name) %>%
  add_tally() %>%
  filter(n > 1) %>%
  mutate(
    first_commit = min(datetime, na.rm = TRUE),
    mid_commit = median(datetime, na.rm = TRUE),
    last_commit = max(datetime, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  geom_linerange(linewidth = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(date_breaks = "2 years", date_labels = "%Y") +
  # coord_flip() +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

@@

We measured the lifespan of a repository as the time between the first and latest commit, and its activity as the rate of commits.
Here therefore refer to the *development* lifespan of a project—measured here as the time between a repository's first and last commit—which is not necessary linked to its use-life.
The lifespan and activity of repositories in open-archaeo vary greatly (@fig-lifespan).
The average project lasts <!-- TODO --> days with <!-- TODO --> commits per day.
<!-- TODO: turn all these percentages into code -->
Many projects are active for only a short period of time: about 15% less than a month, 25% less than three months, and 40% less than a year.
The longest are <!-- TODO -->.
Commit rates vary between <!-- TODO -->, with peaks at <!-- TODO -->.

@@

@@

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\]. A "contribution", in this case, is a commit, issue, or comment applied by a user to a repository.
<!-- TODO: expand -->

```{r plot-contrib-dist}
#TODO: Re-calculate these numbers to ensure they are up to date. Translate the first point to be about how many repos have a quarter of contributions by a single person, so that it corresponds with the vertical lines on the graphic, which are now divided into quarters rather than thirds.
oarch %>%
  unnest(gh_contribs) %>%
  drop_na(contributor) %>%
  group_by(item_name) %>%
  mutate(n_contributors = n()) %>%
  filter(n_contributors > 1) %>%
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) %>% 
  group_by(item_name, big_boy) %>%
  summarise(contributions = sum(contributions)) %>%
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) %>%
  ungroup() %>%
  mutate(item_name = fct_reorder(item_name, p_big_boy)) %>% 
  ggplot(aes(x = item_name, y = contributions,
             fill = fct_relevel(big_boy, c("Other contributors",
                                           "Leading contributor")))) +
  geom_col(position = "fill", colour = "white", size = 0.1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast(guide = guide_legend(reverse = TRUE)) +
  labs(title = "Repository contributors by proportion of contributions",
       x = NULL, y = "Contributions",
       fill = NULL,
       caption = "Data: open-archaeo.info and GitHub API") +
  coord_flip() +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

```{r fig-stars}
#TODO: stars figure?
```

```{r fig-follows}
#TODO: follows figure?
```

Another way GitHub users can engage with repositories and other users with social media-like features such as *starring* a repository, adding it to a public or private list, or *following* a user.
These actions populate a timeline of through which users can see recent activity and discover new projects related to those they have interacted with in the past.^[This feature of GitHub's timeline was one of the primary ways we compiled open-archaeo.]
While not constituting a direct form of collaboration, these features can facilitate the formation and maintenance of collaborative networks, in the same way that other social media platforms serve other professional networks.
